Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2018-10-27 paper/paper.Rmd
Edits to paper from thesis and evaluation results

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/fig_overview.R')
#read_chunk('fig_annotationAndModel.R')
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
invisible(dev.off())
```

## Abstract {-}

Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate current knowledge on these efforts including the open-access Clinical Interpretations of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from highly-skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant markers from all available published literature. A group of cancer genomics experts annotated biomarkers discussed in 800 sentences and achieved good inter-annotator agreement. We used a supervised learning approach to construct the CIViCmine knowledgebase and extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers.  CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

## Author Summary {-}

The future of cancer treatment will involve examining the DNA and other molecular biomarkers of individual patient’s tumours. This will allow for a personalized treatment strategy to be designed for each patient. Current approaches for identifying biomarker-guided treatments require researchers to read vast amounts of scientific literature about cancer mutations. These papers discuss how molecular data can help diagnose a cancer, forecast its aggressiveness, understand the risk of developing cancer and predict whether a drug will be effective. The CIViC database is an online resource collating this information into a central repository. However there are too many papers published every month on this topic to allow for constant manual review and update. To help the curators of CIViC know which papers are relevant, we have built a text mining system that extracted and prioritized `r paper.sentenceCount` sentences discussing cancer mutations. This resource is valuable to the CIViC curators and other researchers working towards personalized cancer treatment.

## Introduction {-}

@@

Several groups have built their own in-house knowledgebases which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing].  To encourage a collaborative effort, the CIViC database (https://civicdb.org) was launched to provide a wiki-like editable online resource where edits and additions are moderated by experts in order to maintain high quality [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of biomarkers (also known as evidence types).

Diagnostic evidence items describe variants that can help a clinician diagnose or exclude a cancer as a particular type or subtype. For instance, the JAK2 V617F mutation is a major diagnostic criterion for myeloproliferative neoplasms to identify polycythemia vera, essential thrombocythemia and primary myelofibrosis. Predictive evidence items describe variants that help predict drug sensitivity or response and are valuable in deciding further treatments. Predictive evidence items often explain mechanisms of resistance in patients who progressed on a drug treatment. For example, the ABL1 T315I missense mutation in the BCR-ABL fusion, predicts poor response to imatinib, a tyrosine kinase inhibitor that would otherwise effectively target BCR-ABL, in patients with chronic myeloid leukemia. Predisposing evidence items describe germline variants that increase the likelihood of developing a particular cancer, such as BRCA1 mutations for breast/ovarian cancer or RB1 mutations for retinoblastoma. Lastly, prognostic evidence items describe variants that predict survival outcome. As an example, colorectal cancers that harbor a KRAS mutation are predicted to have worse survival.

@@

Several other groups have created knowledgebases to aid clinical interpretation of cancer genomes. Many of these projects have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) to coordinate these efforts and have created a federated search mechanism to allow easier analysis across multiple knowledgebases (manuscript in preparation). The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledgebase [@huang2017cancer], Molecular Match, JAX-Clinical Knowledgebase [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance there were only 5 publications with BRAF V600E in title or abstract in PubMed in 2004 compared to 454 citations in 2017. In order to maintain a high quality and up-to-date knowledgebase, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is immensely important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.

@@

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the mirTex knowledgebase, which collates microRNA and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases.

The main objective of this project was to identify frequently discussed cancer biomarkers which fit the CIViC model but are not yet included in the CIViC knowledgebase. We developed an IE-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/), and examples of its use to curate information into the CIViC database. This main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` biomarkers, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for biomarkers not yet included in any other resource.

## Methods {-}

### Corpora {-}

The full PubMed and Pubmed Central Open Access subset corpora was downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as “Introduction”, “Methods”, “Results” and many synonyms of these. The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. In order to avoid duplications of publications in PMCOA and PubMed, the PMIDs of all documents included in PMCOA were used to filter out abstracts from the PubMed corpus. The update files from PubMed were also processed to identify the latest version of each abstract to process.

### Term Lists {-}

@@

All term lists were filtered with a stopwords list. This was based on the most frequent 5,000 words found in the Corpus of Contempory American English [@davies2009385] as well as custom set of terms. It was then merged with common words that occur as gene names (such as ICE).

@@

### Entity extraction {-}

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. It was changed to Spacy for the improved Python bindings in version 2 for this project. This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files.

@@

Acronyms were also detected where possible by identifying terms in parentheses and checking the term before it, for instance “non-small cell lung carcinoma (NSCLC)”. This was done to remove entity mistakes where possible. The acronym detection method takes the short form (the term in brackets) and iterates backwards through the long form (the term before brackets) looking for potential matches for each letter. If the long form and short form has overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If the long form or short form has an associated ontology ID but the other one doesn’t, they are combined them and assign the associated ontology ID. If both long form and short form have ontology IDs but there is no overlap, the short form is disregarded as the long form has more likelihood of getting the specific term right.

@@

### Sentence selection {-}

@@

### Annotation Platform {-}

@@

(ref:annotationplatform) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown would tagged using "Predictve/Prognostic" as it describes a prognostic marker.

@@

### Annotation {-}

@@

For the annotation stage (outlined in Fig \@ref(fig:annotationOverview)), an equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types the prognostic type dominated the other groups. For example, if this step was not done, 100 randomly selected sentences would only contain 2 (on average) from the Diagnostic group. This sampling also provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. All together, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

@@

### Relation extraction {-}

@@

For generation of the knowledgebase, the four evidence type relations were predicted first providing relations including a Gene. The AssociatedVariant relation was then predicted and attached to any existing evidence type relation that included that gene.

### Evaluation {-}

@@

knitr::kable(
  traintest,
  booktabs = TRUE, escape=FALSE,
  caption = 'Number of annotations in the 75%/25% training and test split',
  col.names = c('Annotation','Train #', 'Test #')
  )
```

@@

### Precision-recall Tradeoff {-}

```{r civicmineprcurves, echo=F, eval=T, fig.width=8, fig.asp=1.7,  fig.cap='(ref:civicmineprcurves)'}
grid.arrange(fig_prcurves)
```

@@

### Application to Pubmed and PMCOA {-}

With the thresholds selected, the final models were applied to all sentences extracted from Pubmed and PMCOA. This is a reasonably large computational problem and was tasked to the compute cluster at the Genome Sciences Centre.

In order to manage this compute and provide infrastructure for easy updating with new publications in Pubmed and PMCOA, we made use of the updated Pubrunner infrastructure (paper in preparation - https://github.com/jakelever/pubrunner). This allows for easy distribution of the work across a compute cluster. The resulting data was then pushed to Zenodo (https://zenodo.org/) for perpetual and public hosting. It is stored as a simple tab-delimited file with one row per extracted biomarker in the literature. The data is released with a Creative Commons Public Domain (CC0) license so that other groups can easily make use of it.

### CIViC Matching {-}

In order to make comparisons with CIViC, we downloaded the nightly data file from CIViC (https://civicdb.org/releases) and matched evidence items against each other. The evidence type and IDs for genes and cancers were used for matching. Direct string matching was used to compare drug names for Prognostic biomarkers.  The exact variant was not used for comparison in order to find a genes that contain any biomarkers that match between the two resources.

Some mismatches would with drug names. For example, CIViCmine may capture information about the drug family while CIViC contains information on specific drugs, or a list of drugs. Another challenge with matching with CIViCmine is related to the similarity of cancer types in the Disease Ontology. There are several pairs of similar cancers types that are used interchangably by some researchers and not by others, e.g. stomach cancer and stomach carcinoma. CIViC may contain a biomarker for stomach cancer and CIViCmine matches all the other details except it relates it to stomach carcinoma.

### User interface {-}

@@

## Results {-}

@@

(ref:evidencetypes) The entirety of Pubmed and Pubmed Central Open Access subset were processed to extract the four different evidence types shown.

```{r examplesentences, echo=F, eval=T, fig.width=8, fig.asp=0.7,  fig.cap='(ref:examplesentences)'}
exampleSentences <- read.table('civicmine/exampleSentences.tsv',sep='\t',header=T)
#exampleSentences$Sentence <- gsub(";",'\n',exampleSentences$Sentence)

knitr::kable(
  exampleSentences,
  booktabs = TRUE, escape=FALSE,
  caption = 'Four example sentences for the four evidence types extracted by CIViCmine. The associated Pubmed IDs are also shown for reference.',
  col.names = c('Evidence Type','PubMed ID', 'Sentence')
  )
```

From the full Pubmed and PMCOA corpus, we extracted `r paper.biomarkerCount` biomarkers with a breakdown into the four types (Fig 7). As expected, there are many more Prognostic evidence items than the other three types. Table \@ref(tab:examplesentences) outlines examples of all four of these evidence types. `r paper.percSentencesWithMultiple`% of sentences (`r paper.multiplePerSentence`/`r paper.sentenceCount`) contain more than one evidence item, such as the Predictive example which relates EGFR as a predictive marker in NSCLC to both erlotinib and gefitinib. In total, we extracted `r paper.mentionCount` mentions of biomarkers from `r paper.paperCount` unique papers. These biomarkers relate to `r paper.geneCount` genes, `r paper.cancerCount` cancer types and `r paper.drugCount` drugs.

@@

Lastly, we see that expression related biomarkers dominate the variant types (Fig \@ref(fig:entitycounts)d). Markers based on expression are more likely to be prognostic than those using non-expression data (`r paper.expressionPrognosticPerc`% versus `r paper.nonexpressionPrognosticPerc`%). . The easiest method to explore the importance of a gene in a cancer type is to correlate expression levels with patient survival. With the accessibility of large transcriptome sets and survival data (e.g. TCGA), such assertions have become very common. The ‘mutation’ variant type has a more even split across the four evidence types. The mutation term covers very general phrasing without a specific mention of the actual mutation. The substitution variant type does capture this information but there are far fewer. This reflects the challenge of extracting all the evidence item information from a single sentence. It is more likely for an author to define a mutation in another sentence and then use a general term (e.g. EGFR mutations) when discussing its clinical relevance. There are also a substantial number of evidence items where the variant cannot be identified and are flagged as ‘[unknown]’. These are still valuable but may require more in-depth curation in order to tease out the actual variant.

`r paper.multiCitationPerc`% of the biomarkers extracted (`r paper.multiCitationCount`/ `r paper.biomarkerCount`) are supported by more than one citation. In fact, the most cited biomarker is BRCA1 mutation as a predisposing marker in breast cancer with `r paper.predisposing_brca1_breastcancer_mutation` different papers discussing this. The initial priority for CIViC annotation is on highly cited biomarkers that have not yet been curated into CIViC, in order to eliminate obvious information gaps. However, the single citations may also represent valuable information for precision cancer analysts and CIViC curators focused on specific genes or diseases.

@@

### Use Cases {-}

@@

Knowledgebase curation use case: The main purpose of this tool is to assist in curation of new biomarkers in CIViC. A CIViC curator, looking for a frequently discussed biomarker, would access the CIViCmine Shiny app through a web browser. This would present the table, pie charts and filter options on the left. They would initially filter the CIViCmine results for those not already in CIViC. If they had a particular focus, they may filter by Evidence Type. For example, some CIViC curators may be more interested in Diagnostic, Predictive and Prognostic biomarkers than Predisposing. This is due to the focus on somatic events in many cancer types. They would then look at the table of biomarkers, already sorted by citation count in descending order, and select the top one. This would then populate a table further down the page. Assuming that this is a frequently cited biomarker, there would be many sentences discussing it, which would quickly give the curator a broad view of whether it is accepted in the community. They would then open multiple tabs on their web browser to start looking at several of the papers discussing it. They might select an older paper, close to when it was first established as a biomarker, and a more recent paper from a high-impact journal to gauge the current view of the biomarker. Several of the sentences may obviously cite other papers as being important to establishing this biomarker. The curator would look at these papers in particular, as they may be the most appropriate to curate. Importantly, the curator may want the primary literature source(s), which includes the experimental data supporting this biomarker.

@@

The CIViCmine resource was evaluated by a CIViC curator. They selected the top 100 biomarkers of each evidence item deemed to not already be in CIViC. They rated each one based on its appropriateness for CIViC, whether the text extraction was correct and whether it should be curated in CIViC. Table X shows that…

## Discussion {-}

@@

Apart from the advantages outlined previously, several other factors lead to the decision to use a supervised learning approach to build this knowledgebase. The CIViC knowledgebase could have been used as training data in some form. The papers already in CIViC could have been searched for the sentences discussing the relevant biomarker, which could then have been used to train a supervised relation extraction system. An alternative approach to this problem would have been to use a distant supervision method using the CIViC knowledgebase as seed data. This approach was taken by Peng et al who also attempted to extract relations across sentence boundaries [@peng2017cross]. They chose to focus only on point mutations and extracted 530 within sentence biomarkers and 1,461 cross-sentence biomarkers. These numbers are drastically smaller than the 70,655 extracted in CIViCmine.

>>>>

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/fig_overview.R')
#read_chunk('fig_annotationAndModel.R')
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
source('civicmine/fig_evaluationresults.R')
invisible(dev.off())
```

## Abstract

Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate current knowledge on these efforts. These include the open-access Clinical Interpretations of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from highly-skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant markers from all available published literature. A group of cancer genomics experts annotated biomarkers discussed in 800 sentences and achieved good inter-annotator agreement. We used a supervised learning approach to construct the CIViCmine knowledgebase and extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers.  CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

## Introduction

@@

Several groups have built their own in-house knowledge bases which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing].  To encourage a collaborative effort, the CIViC database (https://civicdb.org) was launched to provide a wiki-like editable online resource where edits and additions are moderated by experts in order to maintain high quality [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of biomarkers (also known as evidence types).

Diagnostic evidence items describe variants that can help a clinician diagnose or exclude a cancer. For instance, the JAK2 V617F mutation is a major diagnostic criterion for myeloproliferative neoplasms to identify polycythemia vera, essential thrombocythemia and primary myelofibrosis. Predictive evidence items describe variants that help predict drug sensitivity or response and are valuable in deciding further treatments. Predictive evidence items often explain mechanisms of resistance in patients who progressed on a drug treatment. For example, the ABL1 T315I missense mutation in the BCR-ABL fusion, predicts poor response to imatinib, a tyrosine kinase inhibitor that would otherwise effectively target BCR-ABL, in patients with chronic myeloid leukemia. Predisposing evidence items describe germline variants that increase the likelihood of developing a particular cancer, such as BRCA1 mutations for breast/ovarian cancer or RB1 mutations for retinoblastoma. Lastly, prognostic evidence items describe variants that predict survival outcome. As an example, colorectal cancers that harbor a KRAS mutation are predicted to have worse survival.

@@

Several other groups have created knowledge bases to aid clinical interpretation of cancer genomes. Many of these projects have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) to coordinate these efforts and have created a federated search mechanism to allow easier analysis across multiple knowledge bases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance there were only 5 publications with BRAF V600E in title or abstract in PubMed in 2004 compared to 454 citations in 2017. In order to maintain a high quality and up-to-date knowledge base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is immensely important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledge base. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledge base.

@@

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledge base. IE methods are also used for automated knowledge base population without a manual curation step. For example, the mirTex knowledge base, which collates microRNA and their targets, uses automated relation extraction methods to populate the knowledge base [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledge bases.

The main objective of this project was to identify frequently discussed cancer biomarkers which fit the CIViC model but are not yet included in the CIViC knowledge base. We developed an IE-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledge base. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). This main contributions of this work are an approach for knowledge base construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` biomarkers, is valuable to all cancer knowledge bases to aid their curation and also as a tool for precision cancer analysts searching for biomarkers not yet included in any other resource.

## Methods

### Corpora

The full PubMed and PubMed Central Open Access subset corpora was downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as “Introduction”, “Methods”, “Results” and many synonyms of these. The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. In order to avoid duplications of publications in PMCOA and PubMed, the PMIDs of all documents included in PMCOA were used to filter out abstracts from the PubMed corpus. The update files from PubMed were also processed to identify the latest version of each abstract to process.

### Term Lists

@@

All term lists were filtered with a stopwords list. This was based on the stopword list from the Natural Language Toolkit [@bird2006nltk] and the most frequent 5,000 words found in the Corpus of Contempory American English [@davies2009385] as well as custom set of terms. It was then merged with common words that occur as gene names (such as ICE).

@@

### Entity extraction

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. It was changed to Spacy [@honnibaljohnson] for the improved Python bindings in version 2 for this project. This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files.

@@

Acronyms were also detected where possible by identifying terms in parentheses and checking the term before it, for instance “non-small cell lung carcinoma (NSCLC)”. This was done to remove entity mistakes where possible. The acronym detection method takes the short form (the term in brackets) and iterates backwards through the long form (the term before brackets) looking for potential matches for each letter. If the long form and short form has overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If only one of the long form or short form has an associated ontology ID, they are combined and assigned the associated ontology ID. If both long form and short form have ontology IDs but there is no overlap, the short form is disregarded as the long form has more likelihood of getting the specific term correct.

@@

### Sentence selection

@@

### Annotation Platform

@@

(ref:annotationplatform) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown would be tagged using "Predictve/Prognostic" as it describes a prognostic marker.

@@

### Annotation

@@

For the annotation stage (outlined in Fig \@ref(fig:annotationOverview)), an equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types as the prognostic type dominated the other groups. If this step was not done, 100 randomly selected sentences would only contain 2 (on average) from the Diagnostic group. However, this sampling provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. All together, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

@@

### Relation extraction

@@

For generation of the knowledge base, the four evidence type relations were predicted first which provided relations including a Gene. The AssociatedVariant relation was then predicted and attached to any existing evidence type relation that included that gene.

### Evaluation

@@

knitr::kable(
  traintest,
  booktabs = TRUE, 
  caption = 'Number of annotations in the training and test sets',
  col.names = c('Annotation','Train', 'Test')
  )
```

@@

### Precision-recall Tradeoff

```{r civicmineprcurves, echo=F, eval=T, fig.width=8, fig.asp=1.2,  fig.cap='(ref:civicmineprcurves)'}
grid.arrange(fig_prcurves)
```

@@

### Application to PubMed and PMCOA

With the thresholds selected, the final models were applied to all sentences extracted from PubMed and PMCOA. This is a reasonably large computational problem and was tasked to the compute cluster at the Genome Sciences Centre.

In order to manage this compute and provide infrastructure for easy updating with new publications in PubMed and PMCOA, we made use of the updated Pubrunner infrastructure (paper in preparation - https://github.com/jakelever/pubrunner). This allows for easy distribution of the work across a compute cluster. The resulting data was then pushed to Zenodo (https://zenodo.org/) for perpetual and public hosting. The data is released with a Creative Commons Public Domain (CC0) license so that other groups can easily make use of it.

### CIViC Matching

In order to make comparisons with CIViC, we downloaded the nightly data file from CIViC (https://civicdb.org/releases) and matched evidence items against each other. The evidence type and IDs for genes and cancers were used for matching. Direct string matching was used to compare drug names for Predictive biomarkers.  The exact variant was not used for comparison in order to find a genes that contain any biomarkers that match between the two resources.

Some mismatches occurred with drug names. For example, CIViCmine may capture information about the drug family while CIViC contains information on specific drugs, or a list of drugs. Another challenge with matching with CIViCmine is related to the similarity of cancer types in the Disease Ontology. There are several pairs of similar cancers types that are used interchangably by some researchers and not by others, e.g. stomach cancer and stomach carcinoma. CIViC may contain a biomarker for stomach cancer and CIViCmine matches all the other details except it relates it to stomach carcinoma.

### User interface

@@

## Results

@@

(ref:evidencetypes) The entirety of PubMed and PubMed Central Open Access subset were processed to extract the four different evidence types shown.

```{r examplesentences, echo=F, eval=T}
exampleSentences <- read.table('civicmine/exampleSentences.tsv',sep='\t',header=T)
exampleSentences$Sentence <- gsub(";",'\n',exampleSentences$Sentence)
exampleSentences$Evidence.Type <- as.character(exampleSentences$Evidence.Type)

suppressWarnings(exampleSentences %>%
  dplyr::mutate_all(linebreak) %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Four example sentences for the four evidence types extracted by CIViCmine. The associated PubMed IDs are also shown for reference.',
    col.names = c('Type','PMID', 'Sentence'),
    row.names = NA
    ) %>%
  kable_styling())
```

From the full PubMed and PMCOA corpus, we extracted `r paper.biomarkerCount` biomarkers with a breakdown into the four types (Figure \@ref(fig:evidencetypes)). As expected, there are many more Prognostic evidence items than the other three types. Table \@ref(tab:examplesentences) outlines examples of all four of these evidence types. `r paper.percSentencesWithMultiple`% of sentences (`r paper.multiplePerSentence`/`r paper.sentenceCount`) contain more than one evidence item, such as the Predictive example which relates EGFR as a predictive marker in NSCLC to both erlotinib and gefitinib. In total, we extracted `r paper.mentionCount` mentions of biomarkers from `r paper.paperCount` unique papers. These biomarkers relate to `r paper.geneCount` genes, `r paper.cancerCount` cancer types and `r paper.drugCount` drugs.

@@

Lastly, we see that expression related biomarkers dominate the variant types (Fig \@ref(fig:entitycounts)d). Markers based on expression are more likely to be prognostic than those using non-expression data (`r paper.expressionPrognosticPerc`% versus `r paper.nonexpressionPrognosticPerc`%). The easiest method to explore the importance of a gene in a cancer type is to correlate expression levels with patient survival. With the accessibility of large transcriptome sets and survival data (e.g. TCGA), such assertions have become very common. The ‘mutation’ variant type has a more even split across the four evidence types. The mutation term covers very general phrasing without a specific mention of the actual mutation. The substitution variant type does capture this information but there are far fewer. This reflects the challenge of extracting all the evidence item information from a single sentence. It is more likely for an author to define a mutation in another sentence and then use a general term (e.g. EGFR mutations) when discussing its clinical relevance. There are also a substantial number of evidence items where the variant cannot be identified and are flagged as ‘[unknown]’. These are still valuable but may require more in-depth curation in order to tease out the actual variant.

Of all the biomarkers extracted, `r paper.multiCitationPerc`% (`r paper.multiCitationCount`/ `r paper.biomarkerCount`) are supported by more than one citation. In fact, the most cited biomarker is BRCA1 mutation as a predisposing marker in breast cancer with `r paper.predisposing_brca1_breastcancer_mutation` different papers discussing this. The initial priority for CIViC annotation is on highly cited biomarkers that have not yet been curated into CIViC, in order to eliminate obvious information gaps. However, the single citations may also represent valuable information for precision cancer analysts and CIViC curators focused on specific genes or diseases.

@@

### Use Cases

@@

Knowledge base curation use case: The main purpose of this tool is to assist in curation of new biomarkers in CIViC. A CIViC curator, looking for a frequently discussed biomarker, would access the CIViCmine Shiny app through a web browser. This would present the table, pie charts and filter options on the left. They would initially filter the CIViCmine results for those not already in CIViC. If they had a particular focus, they may filter by Evidence Type. For example, some CIViC curators may be more interested in Diagnostic, Predictive and Prognostic biomarkers than Predisposing. This is due to the focus on somatic events in many cancer types. They would then look at the table of biomarkers, already sorted by citation count in descending order, and select the top one. This would then populate a table further down the page. Assuming that this is a frequently cited biomarker, there would be many sentences discussing it, which would quickly give the curator a broad view of whether it is accepted in the community. They would then open multiple tabs on their web browser to start looking at several of the papers discussing it. They might select an older paper, close to when it was first established as a biomarker, and a more recent paper from a high-impact journal to gauge the current view of the biomarker. Several of the sentences may obviously cite other papers as being important to establishing this biomarker. The curator would look at these papers in particular, as they may be the most appropriate to curate. Importantly, the curator may want the primary literature source(s), which includes the experimental data supporting this biomarker.

@@

```{r evaluationresults, echo=F, eval=T, fig.width=8, fig.asp=1.0,  fig.cap='(ref:evaluationresults)'}
grid.arrange(fig_evaluationresults)
```

(ref:evaluationresults) The top results in CIViCmine were evaluated by a CIViC curator and measured for three categories (correctness, usability and need). Percentages are shown for each metric and evidence type for Yes, Intermediate and No.

In order to evaluate the curation value of the data provided by CIViCmine, a CIViC curator evaluated the top biomarkers identified by CIViCmine that were not found in CIViC. Biomarkers with high citation counts were selected for each evidence type and filtered to include specific variant information. They were then evaluated for correctness (whether the sentences matched the extracted structured data), usability (whether there was enough information for curation into CIViC) and need (whether this information was lacking in CIViC). Each biomarkers was marked in all three categories with Yes, Intermediate and No. Intermediate scores are used to identify cases where additional information (e.g. reading the full paper or its citations) was needed. Figure \@ref(fig:evaluationresults) shows the summary of the results as percentages for each of the three metrics across the four evidence types. Overall, the results are very positive with 73% of evaluated biomarkers being deemed needed by CIViC. The Predictive evidence item was found to have a larger proportion of unneeded evidence items. This was due to the catch all groups (e.g. EGFR inhibitors) that were deemed to be too vague for inclusion into CIViC but might provide valuable information for other clinical researchers. The high percentage of Intermediate for the usability of Predisposing biomarkers was due to the general variant terms identified (such as Mutation) where the exact variant was unclear and further curation would be needed. Overall these results show that CIViCmine provides valuable data that can be curated into CIViC.

## Discussion

@@

Apart from the advantages outlined previously, several other factors lead to the decision to use a supervised learning approach to build this knowledge base. The CIViC knowledge base could have been used as training data in some form. The papers already in CIViC could have been searched for the sentences discussing the relevant biomarker, which could then have been used to train a supervised relation extraction system. An alternative approach to this problem would have been to use a distant supervision method using the CIViC knowledge base as seed data. This approach was taken by Peng et al who also attempted to extract relations across sentence boundaries [@peng2017cross]. They chose to focus only on point mutations and extracted 530 within sentence biomarkers and 1,461 cross-sentence biomarkers. These numbers are drastically smaller than the 70,655 extracted in CIViCmine.

The reason to not use the CIViC knowledge base in the creation of the training data was taken to avoid any curator-specific bias that may have formed in the selection of papers and biomarkers to curate. This was key to providing a broad and unbiased view of the biomarkers discussed in the literature. CIViC evidence items include additional information such as directionality of a relationship (e.g. does a mutation cause drug sensitivity or resistance), the level of support for it (from preclinical models up to FDA guidelines) and several other factors. It is highly unlikely that all this information will be included within a single sentence. Therefore, we did not try to extract this information concurrently. Instead, it is an additional task for the curator as they process the CIViCmine prioritised list.
